---
title: "Data science project plan"
subtitle: "Example submission"
author: "Student ID: 123456"
toc: false
# # Uncomment the next line to use the default LaTeX template:
format: pdf
execute: 
  cache: true
  eval: false
number-sections: false
bibliography: references.bib
---


**Note: this is an example submission for illustrative purposes only. You are welcome to choose a related topic but ensure your submission is original.**

See the source code at [github.com/itsleeds/tds/tree/main/d2](https://github.com/itsleeds/tds/tree/main/d2/example.qmd).

See the rendered PDF at [gitub.com/itsleeds/tds/releases/](https://github.com/itsleeds/tds/releases/download/2025/project-plan-example.pdf).

See the .zip file with the files needed to reproduce this analysis at [gitub.com/itsleeds/tds/releases/](https://github.com/itsleeds/tds/releases/download/2025/formative-assessment-example.zip).

Note: you need to install LaTeX for the PDF to render, e.g. by running the following command:

```
quarto install tinytex
```

# Introduction

Road traffic casualties cause untold pain and suffering, killing an estimated 1.35 million people worldwide each year and ruining the lives of countless others [@worldhealthorganization2018].
The UK has comparatively safe roads but road traffic casualty rates have plateaued in recent years, according to data from the [Department for Transport's 2023 report](https://www.gov.uk/government/statistics/reported-road-casualties-great-britain-annual-report-2023/reported-road-casualties-great-britain-annual-report-2023).
There are many causes of traffic collisions but road traffic infrastructure can play a role, as outlined in @tait_contraflows_2023, which found that contraflow cycling on one-way streets can improve the cycling experience and increase participation without increasing crash rates.
The topic of this project is to investigate the relationship between road traffic casualties and road infrastructure in West Yorkshire, a region that has been found in previous research to have a higher casualty rate for cyclists than other regions [@lovelace_who_2016].

## Working title

**What are the links between new infrastructure and traffic collisions in West Yorkshire?**

# Data

I plan to use the following datasets:

- STATS19 data from the `stats19` R package, 2014-2023
  - I plan to use the crash-level data converted into spatial format with the `format_sf()` function
- Traffic calming data from the `osmactive` package
  - This dataset from OSM contains information on speed bumps, chicanes, and other traffic calming measures

# Research question

**How does the presence of traffic calming measures correlate with road traffic casualties in West Yorkshire?**

# Initial analysis

I loaded the tidyverse and the `stats19` packages (code not shown) and created a map of road traffic casualties in West Yorkshire by converting the dataset into `sf` format using the `ggplot2` package with the folliwng commands:

```{r}
#| include: false
#| label: setup
pkgs = c("tidyverse", "stats19", "nptscot/osmactive")
# Install pak pkg if not already installed
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
pak::pkg_install(pkgs)
library(tidyverse)
library(stats19)
library(osmactive)
# Set ggplot2 theme
theme_set(theme_minimal())
```


```{r}
#| include: false
#| label: get-data
osm_transport_network = osmactive::get_travel_network("West Yorkshire")
osm_transport_network_20mph = osm_transport_network |>
  filter(maxspeed == "20 mph")
collisions_2023 = stats19::get_stats19(year = 2023, type = "collision")
```

```{r}
#| label: prep-data
class(collisions_2023)
collisions_2023_sf = stats19::format_sf(collisions_2023)
class(collisions_2023_sf)
names(collisions_2023_sf)[1:9]
head(table(collisions_2023_sf$police_force))
collisions_west_yorkshire_sf = collisions_2023_sf |>
  filter(police_force == "West Yorkshire") |>
  # Arrange in descending order of accident severity
  # so most severe accidents are plotted last:
  arrange(desc(accident_severity))
sf::write_sf(collisions_west_yorkshire_sf, "wy.gpkg")
```


```{r}
#| label: plot
#| layout-ncol: 2
osm_transport_network_20mph |>
  ggplot() +
  geom_sf()
ggplot() +
  geom_sf(
    data = collisions_west_yorkshire_sf,
    aes(colour = accident_severity, alpha = accident_severity)
  ) +
  scale_alpha_manual(values = c(0.8, 0.4, 0.2))
```

Road infrastructure data was obtained from the `osmactive` package.

I rendered this document to a PDF file with the following command:

```{r}
#| eval: false
quarto::quarto_render(
  "project-plan.qmd",
  output_format = "pdf",
  output_file = "project-plan.pdf"
)
```

I created a zip file with the files needed to reproduce this analysis with the following command:

```{r}
#| eval: false
zip(
  zipfile = "submission.zip",
  files = c("project-plan.qmd", "project-plan.pdf", "wy.gpkg")
)
```

```{r}
#| eval: false
#| echo: false
quarto::quarto_render(
  "d2/example.qmd",
  output_format = "pdf",
  output_file = "project-plan-example.pdf"
)
system("gh release upload 2025 docs/project-plan-example.pdf --clobber")
file.copy("docs/project-plan-example.pdf", "d2/project-plan-example.pdf")
file.rename("wy.gpkg", "d2/wy.gpkg")
file.remove("formative-assessment-example.zip")
setwd("d2")
zip(
  zipfile = "formative-assessment-example.zip",
  files = c("project-plan-example.pdf", "wy.gpkg", "example.qmd")
)
setwd("..")
file.rename("d2/formative-assessment-example.zip", "formative-assessment-example.zip")
fs::file_size("formative-assessment-example.zip")
system("gh release upload 2025 formative-assessment-example.zip --clobber")
```

# Questions

- How do I get other types of transport infrastructure, such as cycle lanes and traffic calming measures? I have read the documentation at  [nptscot.github.io/osmactive](https://nptscot.github.io/osmactive/) but I am unsure how to proceed.

# References