---
title: "Seminar 1 - Mini-workshop"
bibliography: ../tds.bib
toc: true
execute: 
  cache: true
  message: false
  warning: false
editor: 
  markdown: 
    wrap: 72
format:
  html:
    code-overflow: wrap
---

The best way to learn is by exploring data and answering your own
questions. Here are some datasets that can help you investigate
questions like:

-   What is the Average Daily People/Bikes/Passengers/Cars?

-   What is the typical daily/weekly/monthly demand profile?

-   Where are the points with the highest demand/flows?

# Some interesting datasets ...

Let's explore some interesting datasets. First we will install (if
necessary) and load the packages for this examples

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) install.packages("remotes")
pkgs = c(
    "sf",
    "tidyverse",
    "osmextract",
    "tmap",
    "maptiles"
)
remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)
```

## Motorised vehicles counts: Leeds

Many cities/countries publish data from permanent traffic counters e.g.
ANPR cameras, induction loops or low-cost sensors. We are going to use
data from the sensors in Leeds (available in [Data Mill
North](https://datamillnorth.org/dataset/e6q0n/leeds-annual-traffic-growth))

```{r}
leeds_car_location <- read_csv(
  "https://datamillnorth.org/download/e6q0n/9bc51361-d98e-47d3-9963-aeeca3fa0afc/Camera%20Locations.csv"
  ) 

leeds_car_location_sf <- leeds_car_location |> 
  st_as_sf(coords = c("X","Y"),
           crs = 27700)
```

```{r}
#| echo: false
tm_basemap("OpenStreetMap")+
tm_shape(leeds_car_location_sf)+
  tm_dots("blue")

```

```{r}
leeds_car_2019 <- read_csv(
  "https://datamillnorth.org/download/e6q0n/9e62c1e5-8ba5-4369-9d81-a46c4e23b9fb/Data%202019.csv"
  )
```

If you are interested in open traffic count datasets see
[this](https://juanfonsecals1.github.io/Results_Summary_Maps/)

#### code 

```{r}
leeds_car_2019 |> 
  group_by(Cosit) |> 
  summarise(mean(Volume))


mean_daily_volumes <- leeds_car_2019 |>
  # converting cosit to numeric
  mutate(Cosit = as.numeric(Cosit)) |> 
  # extracting the date
  mutate(time_date = dmy_hm(Sdate),
         # extracts the day
         date = date(time_date)) |> 
  # calculating the total flows for each day
  summarise(Volume = sum(Volume,rm.na = T),
            .by = c(date,Cosit)) |> 
  # Calculating the daily mean 
  summarise(daily_volume = mean(Volume,rm.na = T),
            .by = Cosit) 


daily_volumes <- leeds_car_2019 |> 
    # converting cosit to numeric
  mutate(Cosit = as.numeric(Cosit)) |>
  # extracting the date
  mutate(time_date = dmy_hm(Sdate),
         # extracts the day
         date = date(time_date)) |> # calculating the total flows for each day
  summarise(mean_volume = sum(Volume,rm.na = T),
          .by = c(date,Cosit))

daily_volumes |> 
  mutate(Cosit = as.numeric(Cosit)) |> 
  filter(Cosit == 90201)|> 
  ggplot(aes(x = date,y = mean_volume))+
  geom_line()
  

mean_daily_volumes |> 
  ggplot(aes(daily_volume))+
  geom_histogram()


tmap_mode("view")

leeds_car_location_sf |>
  left_join(daily_volumes,by = c("Site ID"="Cosit")) |> 
  tm_shape()+
  tm_dots("daily_volume")
```

## Cycle counts for West Yorkshire

Some cities would have some dedicated infrastructure to count the number
of people using bikes at strategic points of the city. We are going to
use some cycle counters from West Yorkshire that you can find
[here](https://datamillnorth.org/dataset/e1dmk/leeds-annual-cycle-growth):

```{r}
leeds_bike_location <- read_csv(
  "https://datamillnorth.org/download/e1dmk/a8c8a11e-1616-4915-a897-9ca5ab4e03b8/Cycle%20Counter%20Locations.csv",skip = 1
  ) 

leeds_bike_location_sf <- leeds_bike_location |>
  drop_na(Latitude,Longitude) |> 
  st_as_sf(coords = c("Longitude","Latitude"),
           crs = 4326) |> 
  st_transform(27700)
```

```{r}
#| echo: false
tm_basemap("OpenStreetMap")+
tm_shape(leeds_bike_location_sf)+
  tm_dots("darkblue")
```

The data for 2019:

```{r}
leeds_bike_2019 <- read_csv(
  "https://datamillnorth.org/download/e1dmk/f13f5d49-6128-4619-a3ff-e6e12f88a71f/Cycle%20Data%202019.csv"
  )
```

Other interesting datasets for you to explore are [Paris cycling
counters](https://parisdata.opendatasoft.com/explore/dataset/comptage-velo-donnees-compteurs/dataviz/?disjunctive.id_compteur&disjunctive.nom_compteur&disjunctive.id&disjunctive.name)
or [Scotland](https://usmart.io/org/cyclingscotland/).

## Pedestrian Counts: Melbourne

Cities also monitor the number pedestrians in key locations. We can use
data from the sensors in Melbourne accessible
[here](https://data.melbourne.vic.gov.au/explore/dataset/pedestrian-counting-system-sensor-locations/export/):

```{r}
melbourne_locations_sf <- st_read("https://data.melbourne.vic.gov.au/api/explore/v2.1/catalog/datasets/pedestrian-counting-system-sensor-locations/exports/geojson?lang=en&timezone=Europe%2FLondon")
```

```{r}
#| echo: false
tm_basemap("OpenStreetMap")+
tm_shape(melbourne_locations_sf)+
  tm_dots("darkblue")
```

We will extract

```{r}
melbourne_dec2024 <- read_csv("https://data.melbourne.vic.gov.au/api/explore/v2.1/catalog/datasets/pedestrian-counting-system-monthly-counts-per-hour/exports/csv?lang=en&refine=sensing_date%3A%222024%2F12%22&timezone=Australia%2FMelbourne&use_labels=true&delimiter=%2C")
```

## Public transport tap-in data: Bogotá

Public transport ridership data can be difficult to obtain. Fortunately,
some cities which have systems managed by a public organisation make
this data available for the public. Bogotá's integrated transport system
publishes the tap-in data for the BRT system (see
[this](https://datosabiertos-transmilenio.hub.arcgis.com/)). We will use
one of the daily reports.

```{r}
tm_stations_sf <- st_read("Estaciones_Troncales_de_TRANSMILENIO.geojson")
```

```{r}
#| echo: false
tm_basemap("OpenStreetMap")+
tm_shape(tm_stations_sf)+
  tm_dots("darkblue")
```

Monthly boarding data can be manually obtained in the open data portal
of TransMilenio
[here](https://storage.googleapis.com/validaciones_tmsa/validaciones_mensuales.html?)

```{r}
url_tm <- "https://storage.googleapis.com/validaciones_tmsa/ValidacionTroncal/2024/consolidado_2024.zip"
u_bn <- basename(url_tm)


if(!file.exists(u_bn)){
  download.file(url = url_tm,
                destfile = u_bn,
                mode = "wb")
}



url_tm <- "https://storage.googleapis.com/validaciones_tmsa/ValidacionTroncal/2024/consolidado_2024.zip"


tm_brt_2024 <- read_csv(unz(u_bn,"troncal_2024.csv"))
```

TfL's crowding data is also a great source of ridership data. See
[this](https://datamillnorth.org/download/e1dmk/f13f5d49-6128-4619-a3ff-e6e12f88a71f/Cycle%20Data%202019.csv).

### code

```{r}
daily_tapins <- tm_brt_2024 |> 
  summarise(validaciones = sum(validaciones),
            .by = c(Estacion_Parada,fecha)) |> 
  summarise(validaciones = mean(validaciones),
            .by = Estacion_Parada) |> 
  mutate(numero_estacion = str_extract(Estacion_Parada,"\\(\\d*\\)") |> 
           str_remove_all("(\\(|\\))")) 
  


tm_stations_sf <- st_read("s1/Estaciones_Troncales_de_TRANSMILENIO.geojson")



tm_stations_sf |> 
  left_join(daily_tapins,by = "numero_estacion") |> 
  tm_shape()+
  tm_dots("validaciones",size = "validaciones")
```

## Network data from OSM

You may be already familiar with getting and using OSM data. This an
example of how to obtain the network that can be used for pedestrians.

```{r}
my_coordinates <- c(-76.78893552474851,18.01206727612776)
sf_point <- st_point(my_coordinates) |> st_sfc(crs = 4326)
sf_buffer <- st_buffer(sf_point,dist = 15e3)

```

```{r}
tm_basemap("OpenStreetMap")+
  tm_shape(sf_buffer)+
  tm_borders()
```

```{r}
#| results='hide'
my_network <- oe_get_network(sf_buffer, mode = "walking")
```

```{r}
tm_shape(my_network)+
  tm_lines("highway")
```

***Note:*** you can access a simplified network dataset from Ordnance
Survey's [OpenRoads
dataset](https://www.ordnancesurvey.co.uk/products/os-open-roads).

# Links to other resources

-   See here for a detailed reproducible example of how to visualise
    transport network datasets:
    <https://github.com/juanfonsecaLS1/netvis-trafficflows>
-   See the [`sfnetworks`
    package](https://luukvdmeer.github.io/sfnetworks/) package for
    information on how to work with spatial networks, including
    estimating centality measures and shortest paths.
