---
title: "Coursework submission for Transport Data Science (TRAN5340M)"
subtitle: "Enter your own title here (e.g. Exploring open transport data: a study of the Isle of Wight)"
output: github_document
# output:
#   pdf_document:
#     number_sections: true
author: "Student 12345"
# for html output:
# output:
#   html_document:
#     number_sections: true
# for pdf output:
bibliography: tds.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```

# Introduction

This template contains information and suggested headings for the TDS module.
Do not submit coursework that contains this note or any text (other than the headings) from this template.
It is just designed to get you started. You will submit the .Rmd file, optionally included in a .zip file if you want to include data, and the resulting PDF document as your coursework submission.

As outlined in the module catalogue, the coursework should be:

- A maximum of 3000 words long, excluding code, figure captions and references
- A maximum of 10 pages long, excluding references and appendices (you should include your best work and think about maximising the use of space - the chunk option `out.width="50%"`, for example, can help with this as outlined [here](https://bookdown.org/yihui/bookdown/figures.html) )


# Coursework guidance

The information below provides guidance on the coursework. The fundamentals are:

- Deadline: Friday 15th May, 14:00
- Format: a PDF file (max 10 pages) and an Rmd file or a .zip file containing the Rmd file and minimal dataset needed to reproduce the results if possible (40 MB max size)
- Template: You can download a template .Rmd file as the bais of your submission: https://github.com/ITSLeeds/TDS/raw/master/coursework-template.Rmd

For the coursework you will submit a pdf a document with a maximum of 10 pages that contains code and results demonstrating your transport data science skills.

For an example of the type of submission files you should submit see these:

- A PDF file in an appropriate format https://github.com/ITSLeeds/TDS/releases/download/0.20.1/coursework-template.pdf
- A .zip file containing everything needed to reproduce the results: https://github.com/ITSLeeds/TDS/releases/download/0.20.1/coursework-template.zip

## Marks

Marks are awarded in 5 categories, each of which is evaluated out of 20 and which will take account of the following criteria:

**Understanding of concepts, methods and approaches**

1. Demonstrate understanding of core concepts of data science, including data cleaning/wrangling, joining, exploratory analysis and visualisation.
1. Implement methods appropriate to the objectives of the report.
1. Discuss alternative methods/approaches that could have been used, and consider and their advantages and disadvantages.
1. Describe code that you have written as part of the analysis.

    **Data, processing access and cleaning**

1. Describe how your team obtained and cleaned data
1. Describe methods for cleaning the data and how this added value to it
1. Discuss methods for gaining data using transport techniques such as routing and conversion of origin-destination data to geographic desire lines
2. Describe how your team stored and transported data, including appropriate file formats

    **Application choice and modelling**

1. Choice of topic that has interesting datasets and important real world applications
1. Appropriate use of methods for the choice of application
1. Demonstrate the progression from a simplistic model of the data to modelling techniques
3. Discuss alternative approaches

    **Visualization**

1. Include high quality, attractive and innovative visualisations
2. Describe how they were produced, including code
3. Describe the theory about why they were designed this way
5. Describe how your own work uses concepts from the literature

    **Understand the data science process**

1. The results presented are reproducible using code shared in the report submission or on a code-sharing platform.
1. Evaluation of the computational efficiency of the approach taken and discuss how scalable the code used is and changes that could be made to make it work on larger datasets, e.g. with reference to a benchmark to test different methods/implementations.
1. Provide evidence of teamwork, e.g. how work was divided, scheduled, and managed.
2. Give examples of how you communicated your own ideas to the target audience of the report and within the group.

## Ideas for coursework topics

If you are struggling for ideas and example code, these resources, in addition to the links provided in the lectures and practicals can help:

1. Work through the stats19 training vignette to sharpen your R skills: https://docs.ropensci.org/stats19/articles/stats19-training.html

2. Take a look at the Model Basics chapter (and the next if interested) of the book R for Data Science: https://r4ds.had.co.nz/model-basics.html

3. Try to reproduce the results presented in the ML practical: https://github.com/ITSLeeds/TDS/blob/master/practicals/9-ml.md

When working on the homework you should be thinking about the datasets that you want to use for your coursework assignment. Ideas for datasets that you could use are:

- Travel behaviour data for a region of choice downloaded with the pct package
- Road crash data downloaded with the stats19 package
- OpenStreetMap data downloaded with the osmdata or geofabrik packages

You are free to pick a research topic and research aims and objectives. However, some example topics that you could choose have been provided:

- What explanatory variables best predict the level of walking in Leeds, and how do walking levels relate to pedestrian safety? You could use the command `pct::get_desire_lines(region = "west-yorkshire")` to get data on walking desire lines and `get_stats19()` to get road crash data for Leeds. In terms of the topics covered in the lectures you could:
  - Show understanding of data science in context with a brief introduction that touches on the definition of transport data science
  - Demonstrate understanding of data structures by converting from a data frame of road crashes to an sf object
  - Show routes, e.g. by getting route data with `pct::get_pct_routes_fast()`
  - Show your data visualisation skills by visualising the datasets
  - Demonstrate understanding of modelling with a simple model to explain why the rate of walking varies, e.g. with the distance of the trip and variables that you will calculate (e.g. distance from the city centre)


# Information about RMarkdown

This is an R Markdown file.
You can set the output by changing `output: github_document` to something different, like `output: html_document`.
You will need to submit your work as a pdf document, which can be generated by converting html output to pdf (e.g. with the `pagedown` package) or (recommended) by setting the output to `pdf_document`.
The first lines of your RMarkdown document could look something like this to ensure that the output is a PDF document and that the R code does not run (set `eval = FALSE` to not run the R code):

```
---
title: "Coursework submission for Transport Data Science (TRAN5340M)"
subtitle: "Enter your own title here (e.g. Exploring open transport data: a study of the Isle of Wight)"
output:
  pdf_document:
    number_sections: true
author: "Student 12345"
bibliography: references.bib
---

```

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```



See here for more info: https://rmarkdown.rstudio.com/lesson-2.html


When you open this file in RStudio and click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

To ensure the document is reproducible, you should include a code chunk that shows which packages you used, e.g. to reproduce the results ensure you have installed the necessary packages as follows:

```{r, message=FALSE, eval=FALSE}
# install.packages("remotes")
remotes::install_github("itsleeds/pct")
remotes::install_github("itsleeds/geofabrik")
remotes::install_github("ropensci/stats19")
```

We load the package as follows:

```{r, message=FALSE}
library(pct)
library(sf)
library(stplanr)
library(tidyverse)
library(tmap)
```

You can add references manually or with `[@citation-key]` references linking to a .bib file like this[@lovelace_stplanr_2017].
And this [@fox_data_2018].


## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# Datasets used

You can get zone, OD and even route data for any city in the UK with the following commands.
We got data for the Isle of Wight with the following commands:

```{r, message=FALSE}
library(pct)
region_name = "isle-of-wight"
z = get_pct_zones(region = region_name)
od = get_od()
od_in_zones = od %>% 
  filter(geo_code1 %in% z$geo_code) %>% 
  filter(geo_code2 %in% z$geo_code) 
desire_lines = od2line(od_in_zones, z)
```


You could get data from OpenStreetMap with the `osmdata` package.

```{r, eval=FALSE}
library(osmdata)
osm_data = opq("isle of wight") %>% 
  add_osm_feature(key = "highway", value = "primary") %>% 
  osmdata_sf()
```

```{r, echo=FALSE}
# to use pre-saved data online you can do something like this:
# saveRDS(osm_data, "osm_data_highways_primary.Rds")
# piggyback::pb_upload("osm_data_highways_primary.Rds")
# piggyback::pb_download_url("osm_data_highways_primary.Rds")
u = "https://github.com/ITSLeeds/TDS/releases/download/0.20.1/osm_data_highways_primary.Rds"
osm_data = readRDS(url(u))
```

You can get large OSM datasets with `geofabrik`:

```{r}
library(geofabrik)
iow_highways = get_geofabrik(name = "Isle of Wight", layer = "lines")
summary(as.factor(iow_highways$highway))
iow_highways2 = iow_highways %>% 
  filter(!is.na(highway)) %>% 
  filter(!str_detect(string = highway, pattern = "primary|track|resi|service|foot"))
summary(as.factor(iow_highways2$highway))
```


You could get road casualty data with the `stats19` pakckage, as shown below.

```{r}
crashes = stats19::get_stats19(year = 2018, output_format = "sf") %>% 
  sf::st_transform(crs = sf::st_crs(z))

crashes_in_region = crashes[z, ]
tm_shape(z) +
  tm_fill("car_driver", palette = "Greys") +
  tm_shape(iow_highways2) +
  tm_lines(col = "green", lwd = 2) +
  tm_shape(osm_data$osm_lines) +
  tm_lines(col = "maxspeed", lwd = 5) +
  tm_shape(crashes_in_region) +
  tm_dots(size = 0.5, col = "red")
```

# Descriptive analysis

```{r}
plot(desire_lines)
```

# Route analysis

See [here](https://geocompr.robinlovelace.net/transport.html#routes) and [here](https://www.r-spatial.org/r/2019/09/26/spatial-networks.html) for details.

```{r, echo=FALSE}
# devtools::install_github("luukvdmeer/sfnetworks")
```


```{r}
sln = SpatialLinesNetwork(iow_highways2)
sln_clean = sln_clean_graph(sln)
plot(sln_clean@sl$`_ogr_geometry_`)
centrality = igraph::edge_betweenness(sln_clean@g)
centrality_normalised = centrality / mean(centrality)
```

```{r}
mapview::mapview(z) +
  mapview::mapview(sln_clean@sl, lwd = centrality_normalised * 3, zcol = "maxspeed")
```





# Additional datasets

# Policy analysis

Here you could explain how you explored answers to policy questions such as:

- how to make the roads safer?
- how to reduce congestion?
- where to build bike parking?

# Discussion

Include here limitations and ideas for further research.

# Conclusion

What are the main things we have learned from this project?

To make the code reproducible I saved the code and data as follows


```{r, eval=FALSE}
# Save outputs into zip file:
zip(zipfile = "coursework-template.zip", files = c(
  "coursework-template.Rmd",
  "tds.bib",
  "timetable.csv"
))
# piggyback::pb_upload("coursework-template.zip") # ignore this command
# piggyback::pb_upload("coursework-template.pdf") # ignore this command
```

# References
