---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, eval=FALSE, echo=FALSE}
refs1 = RefManageR::ReadZotero(group = "418217", .params = list(collection = "R38L2EXB", limit = 100))
refs2 = RefManageR::ReadZotero(group = "418217", .params = list(collection = "JFR868KJ", limit = 100))
RefManageR::WriteBib(c(refs1, refs2), "tds.bib")
```


# TDS (Transport Data Science)

This is a GitHub Repository (repo for short) that supports teaching of the Transport Data Science module at the University of Leeds.
The module can be taken by students on the [Data Science and Analytics](https://courses.leeds.ac.uk/i071/data-science-and-analytics-msc) and (from 2022 onwards) [Transport Planning and the Environment](https://courses.leeds.ac.uk/a386/transport-planning-and-the-environment-msc) MSc courses.

The module catalogue can be found at [`catalogue.md`](https://github.com/ITSLeeds/TDS/blob/master/catalogue.md).
The computer code accompanying the course can be found in the `code` folders.
To run this code you will need R and Python installed plus various packages and libraries.
<!-- This software has been packaged-up into a docker container to ease teaching. -->

The timetable can be found:

- On the University's system (official): [http://timetable.leeds.ac.uk/](http://timetable.leeds.ac.uk/teaching/202021/reporting/Individual?objectclass=module&idtype=name&identifier=TRAN5340M01&&template=SWSCUST+module+Individual&days=1-7&weeks=1-52&periods=1-21)
- In ical format (for import into Google/Outlook/other Calendar systems): https://github.com/ITSLeeds/TDS/raw/master/timetable-2020.ics
- As a .csv file (for easy reading as data): https://github.com/ITSLeeds/TDS/blob/master/timetable.csv

<!-- # References -->

<!-- To access references collected for this course (and contribute more if you want), you can join the 'tds' Zotero group: https://www.zotero.org/groups/956304/tds -->

# Software

For this module you need to have up-to-date versions of R and RStudio.
Install:

- R from [cran.r-project.org](https://cran.r-project.org/)
- RStudio from [rstudio.com](https://rstudio.com/products/rstudio/download/#download)
- R packages, by opening RStudio and typing `install.packages("stats19")` in the R console.

We recommend using at least the latest stable release of R (4.0.0 or above). 
We recommend running R on a decent computer, with at least 4 GB RAM and ideally 8 GB or more RAM.
See [Section 1.5 of the online guide Reproducible Road Safety Research with R](https://itsleeds.github.io/rrsrr/introduction.html#installing-r-and-rstudio) for instructions on how to install key packages we will use in the module.^[
For further guidance on setting-up your computer to run R and RStudio for spatial data, see these links, we recommend 
Chapter 2 of Geocomputation with R (the Prerequisites section contains links for installing spatial software on Mac, Linux and Windows): https://geocompr.robinlovelace.net/spatial-class.html and Chapter 2 of the online book *Efficient R Programming*, particularly sections 2.3 and 2.5, for details on R installation and [set-up](https://csgillespie.github.io/efficientR/set-up.html) and the 
[project management section](https://csgillespie.github.io/efficientR/set-up.html#project-management).
]

<!-- ## Course locations -->

<!-- See the image below for the course locations and the following links: -->

<!-- The lectures will be in the Business School Maurice Keyworth SR (1.15): http://students.leeds.ac.uk/room/1-01-087-2730-01-115 -->

<!-- The practicals will be in the West Teaching Lab Cluster (B.16): http://it.leeds.ac.uk/site/custom_scripts/clusters.php -->

# Slides and lectures

Slides can be found online:

- See https://itsleeds.github.io/TDS/slides/1-intro.html#1 for the introductory slides, for example
- Videos of the lectures can be found on the University of Leeds' Blackboard system (you must must [register](https://www.leeds.ac.uk/info/101040/applying/86/how_to_apply_for_masters_courses) to courses such as [Data Science and Analytics](https://courses.leeds.ac.uk/i071/data-science-and-analytics-msc) or [Transport Planning and the Environment](https://courses.leeds.ac.uk/a386/transport-planning-and-the-environment-msc) to take the course)

# Issues and contributing

Any feedback or contributions to this repo are welcome.
If you have a question please open an issue here (you'll need a GitHub account): https://github.com/ITSLeeds/TDS/issues

<!-- # Data -->

<!-- Data for course can be accessed from the repos [Releases](https://github.com/ITSLeeds/TDS/releases) page. -->
<!-- You can, for example, download and unzip the data folder in a local version of the repo (accessed by downloading and unzipp https://github.com/ITSLeeds/TDS/archive/master.zip ) with the following R commands: -->

```{r, eval=FALSE, echo=FALSE}
download.file("https://github.com/ITSLeeds/TDS/releases/download/0.1/data.zip", destfile = "data.zip")
unzip("data.zip")
```

<!-- If you want to be clever you can use the piggyback package: -->

```{r, eval=FALSE, echo=FALSE}
install.packages("piggyback")
piggyback::pb_download("data.zip")

# (This package was used to upload the data with:)
# piggyback::pb_upload(file = "data.zip")
# piggyback::pb_upload(file = "codeExamples.zip")
```



```{r, echo=FALSE, eval=FALSE}
# aim: sort csv file
e = read.csv("sample-data/everyone.csv")
dplyr::arrange(.data = e, n_coffee)
```

```{r, eval=FALSE, engine='python', echo=FALSE}
import pandas as pd
e = pd.read_csv("/mnt/27bfad9a-3474-4e61-9a43-0156ebc67d67/home/robin/ITSLeeds/TDS/sample-data/everyone.csv")
pd.DataFrame.sort_values(e, "n_coffee")
```

```{r, eval=FALSE, echo=FALSE}
# Sorting in the tidyverse
library(tidyverse)
e = read_csv("sample-data/everyone.csv")
e %>% arrange(n_coffee)
e %>% arrange(desc(n_coffee))
```

<!-- # Other projects -->

<!-- - A book on R for Geocomputation: https://github.com/Robinlovelace/geocompr -->
<!-- - A Python package for OSM data analysis: https://github.com/gboeing/osmnx -->

<!-- # Building the website -->

<!-- To publish the slides and other content online, the following commands were used: -->

```{r, echo=FALSE, eval=FALSE}
# first time
# git checkout --orphan gh-pages
# git rm -rf .
# 
# # create a hidden file .nojekyll
# touch .nojekyll
# git add .nojekyll
# 
# git commit -m"Initial commit"
# git push origin gh-pages

# Then...
source deploy.sh
```