---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, eval=FALSE, echo=FALSE}
refs1 = RefManageR::ReadZotero(group = "418217", .params = list(collection = "R38L2EXB", limit = 100))
refs2 = RefManageR::ReadZotero(group = "418217", .params = list(collection = "JFR868KJ", limit = 100))
RefManageR::WriteBib(c(refs1, refs2), "tds.bib")
```


# TDS (Transport Data Science)

This repo supports teaching of the Transport Data Science module at the University of Leeds.

The module catalogue can be found at [`catalogue.md`](https://github.com/ITSLeeds/TDS/blob/master/catalogue.md).
The code accompanying the course can be found in the `code` folders.
To run this code you will need R and Python installed plus various packages and libraries.
This software has been packaged-up into a docker container to ease teaching.

See here for the the timetable, a basic visualisation of which is shown below:

- On the University's system (official): [http://timetable.leeds.ac.uk/](http://webprod3.leeds.ac.uk/catalogue/dynmodules.asp?Y=201920&M=TRAN-5340M)
- In .ics form: https://github.com/ITSLeeds/TDS/releases/download/0.20.1/tds-timetable.ics
- As a .csv file: https://github.com/ITSLeeds/TDS/blob/master/timetable.csv

# References

To access references collected for this course (and contribute more if you want), you can join the 'tds' Zotero group: https://www.zotero.org/groups/956304/tds

# Software

For this module you need to have up-to-date versions of R and RStudio.
For guidance on setting-up your computer to run R and RStudio for spatial data, see these links:

- Chapter 2 of Geocomputation with R (the Prerequisites section contains links for installing spatial software on Mac, Linux and Windows): https://geocompr.robinlovelace.net/spatial-class.html

- Chapter 2 of the online book *Efficient R Programming*, particularly sections 2.3 and 2.5, for details on R installation and [set-up](https://csgillespie.github.io/efficientR/set-up.html) and the 
[project management section](https://csgillespie.github.io/efficientR/set-up.html#project-management).

## Course locations

See the image below for the course locations and the following links:

The lectures will be in the Business School Maurice Keyworth SR (1.15): http://students.leeds.ac.uk/room/1-01-087-2730-01-115

The practicals will be in the West Teaching Lab Cluster (B.16): http://it.leeds.ac.uk/site/custom_scripts/clusters.php

# Issues and contributing

Any feedback or contributions to this repo are welcome.
If you have a question please open an issue here (you'll need a GitHub account): https://github.com/ITSLeeds/TDS/issues

# Data

Data for course can be accessed from the repos [Releases](https://github.com/ITSLeeds/TDS/releases) page.
You can, for example, download and unzip the data folder in a local version of the repo (accessed by downloading and unzipp https://github.com/ITSLeeds/TDS/archive/master.zip ) with the following R commands:

```{r, eval=FALSE}
download.file("https://github.com/ITSLeeds/TDS/releases/download/0.1/data.zip", destfile = "data.zip")
unzip("data.zip")
```

If you want to be clever you can use the piggyback package:

```{r, eval=FALSE}
install.packages("piggyback")
piggyback::pb_download("data.zip")

# (This package was used to upload the data with:)
# piggyback::pb_upload(file = "data.zip")
# piggyback::pb_upload(file = "codeExamples.zip")
```



```{r, echo=FALSE, eval=FALSE}
# aim: sort csv file
e = read.csv("sample-data/everyone.csv")
dplyr::arrange(.data = e, n_coffee)
```

```{r, eval=FALSE, engine='python'}
import pandas as pd
e = pd.read_csv("/mnt/27bfad9a-3474-4e61-9a43-0156ebc67d67/home/robin/ITSLeeds/TDS/sample-data/everyone.csv")
pd.DataFrame.sort_values(e, "n_coffee")
```

```{r, eval=FALSE, echo=FALSE}
# Sorting in the tidyverse
library(tidyverse)
e = read_csv("sample-data/everyone.csv")
e %>% arrange(n_coffee)
e %>% arrange(desc(n_coffee))
```

# Other projects

- A book on R for Geocomputation: https://github.com/Robinlovelace/geocompr
- A Python package for OSM data analysis: https://github.com/gboeing/osmnx

<!-- # Building the website -->

<!-- To publish the slides and other content online, the following commands were used: -->

```{r, echo=FALSE, eval=FALSE}
# first time
# git checkout --orphan gh-pages
# git rm -rf .
# 
# # create a hidden file .nojekyll
# touch .nojekyll
# git add .nojekyll
# 
# git commit -m"Initial commit"
# git push origin gh-pages

# Then...
source deploy.sh
```