---
title: "Schedule"
execute: 
  echo: false
---

<<<<<<< HEAD
<!-- From https://timetable.leeds.ac.uk/teaching/202425/reporting/Individual?objectclass=module&idtype=name&identifier=TRAN5340M01&&template=SWSCUST+module+Individual&days=1-7&weeks=1-52&periods=1-21 -->

The timetable below is a user-friently representation of the timetable for the module (see github for [.csv](https://github.com/itsleeds/tds/blob/main/timetable.csv) and [.ics](https://github.com/itsleeds/tds/releases/download/2025/timetable.ics) versions).
See [timetable.leeds.ac.uk](https://timetable.leeds.ac.uk/teaching/202425/reporting/Individual?objectclass=module&idtype=name&identifier=TRAN5340M01&&template=SWSCUST+module+Individual&days=1-7&weeks=1-52&periods=1-21) and [mytimetable.eeds.ac.uk](https://mytimetable.leeds.ac.uk/link?timetable.id=202425!module!5216C608F8794D77F15FA9D195AB3C06) for the official timetable.
=======
<!-- From https://timetable.leeds.ac.uk/teaching/202526/reporting/Individual?objectclass=module&idtype=name&identifier=TRAN5340M01&&template=SWSCUST+module+Individual&days=1-7&weeks=1-52&periods=1-21 -->

The timetable below is a user-friently representation of the timetable for the module (see github for [.csv](https://github.com/itsleeds/tds/blob/main/data/timetable.csv) and [.ics](https://github.com/itsleeds/tds/releases/download/2026/timetable.ics) versions).
See [timetable.leeds.ac.uk](https://mytimetable.leeds.ac.uk/link?timetable.id=202526!module!5216C608F8794D77F15FA9D195AB3C06) and click on "[2025/26] January to May" in the dropdown menu that says "Month from selected day" for the official timetable.
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
If you spot any discrepancies, please let us know.

```{r}
#| include: false
library(tidyverse)
```

```{r}
#| eval: false
# Aim: create ical of all lectures and practicals of TDS
<<<<<<< HEAD
# Start date of week 1 (source: https://ses.leeds.ac.uk/info/21630/timetabling/1384/teaching-week-patterns-202425)
w_start = as.Date("2024-09-30")
=======
# Start date of week 1 (source: https://www.leeds.ac.uk/term-dates)
w_start = as.Date("2025-09-29")
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
w_start
lubridate::wday(w_start, label = TRUE) # start on a Monday
week_num = c(1:11, paste0("C", 1:4), 12:22, paste0("C", 1:4), 23:30)
n_weeks = length(week_num)
week_commencing = seq(from = w_start, by = 7, length.out = n_weeks)
weeks = tibble::tibble(week_num, week_commencing, day = lubridate::wday(week_commencing, label = TRUE))

<<<<<<< HEAD
# practical sessions ------------------------------------------------------

practical_ids = c(
  "intro",
  "getting",
  "od",
  "routing",
  "visualisation",
  "modelling",
  "project"
)

practical_descriptions = c(
  "Introduction to transport data science (RL, MM, ZW, YY)",
  "Getting transport data (RL, YY)",
  "Origin-destination data (RL, YY)",
  "Routing (MM, YY)",
  "Visualising transport data (RL, YY)",
  "Joining and combining datasets (RL, YY)",
  "Project work (RL, MM, ZW, YY)"
=======
# session timetable ------------------------------------------------------

session_ids = c(
  "intro",
  "getting",
  "ai",
  "od",
  "routing",
  "modelling"
)

session_descriptions = c(
  "Introduction to transport data science (RL, MM, YY)",
  "Getting transport data (RL, YY)",
  "AI for transport planning (RL, YY)",
  "Origin-destination data (RL, YY)",
  "Routing (MM, RL)",
  "Joins, models and publishing your work (RL, MM, YY)"
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
)

# Source: https://timetable.leeds.ac.uk/teaching/202425/reporting/textspreadsheet;?objectclass=module&idtype=name&identifier=TRAN5340M01&&template=SWSCUST+module+Individual&days=1-7&weeks=1-52&periods=1-21
# They happen on Thursdays, 10:00-13:00
<<<<<<< HEAD
practical_day_of_week = 4
practical_start_time = "10:00"
practical_end_time = "13:00"
# From https://timetable.leeds.ac.uk/teaching/202425/reporting/Individual?objectclass=module&idtype=name&identifier=TRAN5340M01&&template=SWSCUST+module+Individual&days=1-7&weeks=1-52&periods=1-21
# Richard Hughes Cluster (1.40) 		23
# Richard Hughes Cluster (1.40) 		15-17, 19, 22
# Richard Hughes Cluster (1.40) 		14	  	 
week_num = as.character(c(14, 15:17, 19, 22, 23))
practical = tibble::tibble(week_num = week_num)
practical = dplyr::inner_join(practical, weeks)
practical$date = practical$week_commencing + (practical_day_of_week - 1)
practical$DTSTART = lubridate::ymd_hm(paste(practical$date, practical_start_time)) 
practical$DTEND = lubridate::ymd_hm(paste(practical$date, practical_end_time))
practical$duration = (practical$DTEND - practical$DTSTART)
practical$type = "Computer practical and lecture"
practical$SUMMARY = paste0("TDS Practical ", 1:nrow(practical), ": ", practical_descriptions)
practical$LOCATION = "Richard Hughes Cluster (1.40)"
# 4th practical is in ITS room 1.11
practical$LOCATION[4] = "Institute for Transport Studies 1.11"
practical$DESCRIPTION = paste0(practical$SUMMARY, " Located in ", practical$LOCATION)
nrow(practical) # 7 practicals
=======
session_day_of_week = 4
session_start_time = "10:00"
session_end_time = "13:00"
# From https://timetable.leeds.ac.uk/teaching/202526/reporting/Individual?objectclass=module&idtype=name&identifier=TRAN5340M01&&template=SWSCUST+module+Individual&days=1-7&weeks=1-52&periods=1-21
# Michael Sadler SR (LG.15) Multi-mode - weeks 14-17, 19, 22
week_num = as.character(c(14, 15:17, 19, 23))
sessions = tibble::tibble(week_num = week_num)
sessions = dplyr::inner_join(sessions, weeks)
sessions$date = sessions$week_commencing + (session_day_of_week - 1)
sessions$DTSTART = lubridate::ymd_hm(paste(sessions$date, session_start_time)) 
sessions$DTEND = lubridate::ymd_hm(paste(sessions$date, session_end_time))
sessions$duration = (sessions$DTEND - sessions$DTSTART)
sessions$type = "Lecture and Seminar"
sessions$SUMMARY = paste0("TDS Session ", 1:nrow(sessions), ": ", session_descriptions)
sessions$LOCATION = "Michael Sadler SR (LG.15)"
# Update Session 4 to ITS Room 1.11a-c (week 17)
sessions$LOCATION[sessions$week_num == "17"] = "Institute for Transport Studies Room 1.11a-c"
sessions$LOCATION[sessions$week_num == "23"] = "Mechanical Engineering LT B (2.37)"
sessions$DESCRIPTION = paste0(sessions$SUMMARY, " Located in ", sessions$LOCATION)
nrow(sessions) # 6 sessions
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33

# seminars ------------------------------------------------------

seminar_ids = c(
  "seminar1",
  "seminar2"
)
seminar_descriptions = c(
<<<<<<< HEAD
  "Seminar 1: Juan Fonseca, University of Leeds: Traffic estimation",
  "Seminar 2 Will Deakin, Network Rail: Network simplication"
)

seminar_day_of_week = c(4)
seminar_start_time = c("14:00", "10:00")
seminar_end_time = c("17:00", "11:00")
=======
  "Seminar 1: Juan Fonseca, University of Leeds: Traffic estimation and transport data visualisation",
  "Seminar 2: Will Deakin, Network Rail: Network simplification"
)

seminar_day_of_week = c(4, 4)
seminar_start_time = c("14:00", "10:00")
seminar_end_time = c("17:00", "13:00")
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
seminar = tibble::tibble(week_num = as.character(c(17, 21)))
seminar = dplyr::inner_join(seminar, weeks)
seminar$date = seminar$week_commencing + (seminar_day_of_week - 1)
seminar$DTSTART = lubridate::ymd_hm(paste(seminar$date, seminar_start_time)) 
seminar$DTEND = lubridate::ymd_hm(paste(seminar$date, seminar_end_time))
seminar$duration = (seminar$DTEND - seminar$DTSTART)
seminar$type = "Seminar"
seminar$SUMMARY = paste0("TDS seminar ", 1:nrow(seminar))
<<<<<<< HEAD
seminar$LOCATION = "Institute for Transport Studies 1.11"
=======
seminar$LOCATION = "Institute for Transport Studies Room 1.11a-c"
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
seminar$DESCRIPTION = paste0(seminar_descriptions, "")
nrow(seminar) # 2 seminars

# deadlines ------------------------------------------------------

deadline_ids = c(
  "computer setup",
  "portfolio draft",
  "portfolio deadline"
)
deadline_descriptions = c(
  "Computer set-up",
  "Draft portfolio",
  "Deadline: coursework, 2pm"
)

# Deadline is 15th May: https://minerva.leeds.ac.uk/webapps/blackboard/content/listContentEditable.jsp?content_id=_629207_1&course_id=_504933_1&mode=reset

deadline_day_of_week = 5
deadline_start_time = "13:00"
deadline_end_time = c("15:01", "13:01", "13:01")
deadline = tibble::tibble(week_num = as.character(c(14, 18, 25)))
deadline = dplyr::inner_join(deadline, weeks)
deadline$date = deadline$week_commencing + (deadline_day_of_week - 1)
deadline$DTSTART = lubridate::ymd_hm(paste(deadline$date, deadline_start_time)) 
deadline$DTEND = lubridate::ymd_hm(paste(deadline$date, deadline_end_time))
deadline$duration = (deadline$DTEND - deadline$DTSTART)
deadline$type = "Deadline"
deadline$SUMMARY = paste0("TDS deadline ", 1:nrow(deadline))
deadline$LOCATION = "Online - Teams"
deadline$DESCRIPTION = deadline_descriptions

<<<<<<< HEAD
# setdiff(names(seminar), names(practical))
timetable = rbind(practical, seminar, deadline) 
=======
# setdiff(names(seminar), names(sessions))
timetable = rbind(sessions, seminar, deadline) 
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
timetable$duration

timetable$UID = purrr::map_chr(1:nrow(timetable), ~ calendar::ic_guid())
timetable = timetable |> 
  arrange(DTSTART) 
units(timetable$duration) = "hours"

sum(timetable$duration) # 20 - up to 25.05 hours of contact time
<<<<<<< HEAD
# xlsx::write.xlsx(timetable, "tds-timetable-2020.xlsx")
=======
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33

ic = calendar::ical(timetable) 
tt_min = dplyr::select(timetable, SUMMARY, DESCRIPTION, DTSTART, DTEND, LOCATION, UID)
ic = calendar::ical(tt_min)
class(tt_min$DTSTART)
format(tt_min$DTSTART, "%Y-%m-%d %H:%M")
tt_csv = tt_min |> 
  mutate(
    time = DTSTART,
    duration = round(as.numeric(DTEND - DTSTART) / 60)
  ) |> 
  select(SUMMARY, DESCRIPTION, time, duration, LOCATION)
tt_csv$time
names(tt_csv) = tolower(names(tt_csv))
<<<<<<< HEAD
calendar::ic_write(ic, "timetable.ics")
readLines("timetable.ics")
readr::write_csv(tt_csv, "timetable.csv")
```

```{r}
#| eval: false
# system("gh release create 2025")
system("gh release upload 2025 timetable.ics timetable.csv --clobber")
=======
calendar::ic_write(ic, "data/timetable.ics")
readLines("data/timetable.ics")
readr::write_csv(tt_csv, "data/timetable.csv")
# # Export timetable to Excel:
# writexl::write_xlsx(tt_csv, "~/OneDrive/career/modules/tds/2026/TRAN5340M-timetable-draft-v2.xlsx")
# system("gh release create 2026")
system("gh release upload 2026 data/timetable.ics data/timetable.csv --clobber")
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
```

```{r}
#| include: false
<<<<<<< HEAD
tt_csv = readr::read_csv("timetable.csv")
=======
tt_csv = readr::read_csv("data/timetable.csv")
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
```


```{r}
timetable = tt_csv |> 
  mutate(
<<<<<<< HEAD
    session_code = paste0(
      # Take 5th character from summary:
      tolower(str_sub(summary, 5, 5)),
      # Take first number from summary:
      str_extract(summary, "\\d")
=======
    session_code = case_when(
      str_detect(summary, "TDS Session") ~ paste0("s", str_extract(summary, "\\d")),
      str_detect(summary, "TDS seminar") ~ paste0("sem", str_extract(summary, "\\d")),
      str_detect(summary, "TDS deadline") ~ paste0("d", str_extract(summary, "\\d")),
      TRUE ~ "unknown"
>>>>>>> 21c60dadd8afab33df597e739a9edfba5d586d33
    )
  ) |>
  transmute(
    `Session ID` = paste0(
      "<a href='",
      session_code,
      "'>",
      session_code,
      "</a>"
    ),
    Description = description,
    Time = format(as.POSIXct(time), "%Y-%m-%d %H:%M"),
    Duration = duration,
    Location = location
  )
DT::datatable(timetable, rownames = FALSE, escape = FALSE, options = list(pageLength = 20))
```

