---
title: "Joins and Aggregations in Transport Data Science"
subtitle: '<br/>🗺<br/>Transport Data Science'
author: "Robin Lovelace"
date: 'University of Leeds'
format: revealjs
bibliography: ../tds.bib
execute: 
  eval: false
  echo: true
---

## Objectives

- Understand different types of joins (spatial and key-based)
- Learn how to perform spatial joins with sf and dplyr
- Apply aggregations to summarize data
- Visualize joined datasets

```{r}
#| label: setup
library(sf)
library(dplyr)
library(tmap)
library(stats19)
```

## What is a Join?

::: incremental
- **Combining datasets** based on common attributes or spatial relationships
- **Spatial joins**: Link points (e.g., crashes) to polygons (e.g., LSOAs) using geometry
- **Key-based joins**: Match IDs across tables (e.g., LSOA codes)
:::

## Spatial Join Example

1. **Load spatial datasets**:
```{r}
path <- "https://github.com/itsleeds/tds/releases/download/2025/p6-lsoa_boundary_wy.geojson"
lsoa <- read_sf(path)
crashes <- stats19::get_stats19("2023")
```

2. **Perform spatial join**:
```{r}
joined <- st_join(lsoa, crashes, join = st_intersects)
```

3. **Aggregate results**:
```{r}
agg_data <- joined |>
  group_by(lsoa_id) |>
  summarize(total_crashes = n())
```

## Key-Based Joins

Use `dplyr` functions like `left_join` to combine data by identifiers:

```{r}
# Example: Join crash counts with population data
final <- left_join(agg_data, pop_data, by = "lsoa_id")
```

## Aggregations

- Group data by spatial units (e.g., LSOA)
- Calculate metrics like **crashes per capita**:

```{r}
final <- final |>
  mutate(crashes_per_capita = total_crashes / population)
```

## Visualization

Create maps to visualize crash patterns using `tmap`:

```{r, warning=FALSE}
tmap_mode("plot")
tm_shape(final) +
  tm_polygons("crashes_per_capita")
```

## References
{references}
