---
title: "Local route network analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Routing from A to B

Routing is the process of finding the "shortest" path from A to B. In this context shortest does not just mean in distance, it may be in time (quickest), or some other characteristic e.g. safest, quietest. 

There are many packages that enable you to do routing in R. When choosing a package you should consider several characteristics:

### Local or Remote

Some packages can do local routing on your own computer. While others allow you to connect to a service.

**Local Routing**

* Usually requires more effort to set up
* No cost (except for time and hardware)
* Control over data, custom scenarios possible futures etc

**Remote Routing**

* Easy setup
* May charge or limit the number of routes
* May support more complex options e.g. traffic, public transport
* Usually limited to routing in the present e.g. current road network current transport timetables.

### Routing Features

Not all routing services can do all types of routing, or do them equally well. Most do driving directions but consider if they do:

* Walking / Cycling (if so does it include specialist road types, exclude dangerous roads)
* Take account of hilliness
* Public Transport (if so does it include fares, which types?)
* Are public transport routes based on timetables or real-time service status?
* Take account of steps and disabled access
* Support specialist vehicles (e.g. lorries and low bridges)
* Does it support real-time or historical traffic data?

## Routing packages for R

A non-comprehensive list of routing packages for R

## Packages on CRAN

* [googleway](https://cran.r-project.org/web/packages/googleway/) support for Google Maps and Directions
* [mapsapi](https://cran.rstudio.com/web/packages/mapsapi/) alternative for google maps
* [osrmr](https://cran.r-project.org/web/packages/osrmr/) Open Source Routing Machine, can connect to remote 
* [CycleStreets](https://cran.r-project.org/web/packages/cyclestreets) Specalist cycling routing, used by 
* [dodgr](https://cran.r-project.org/web/packages/dodgr) Routing done in R
* [igraph](https://cran.r-project.org/web/packages/igraph) General network analysis, not transport specific
* [stplanr](https://cran.r-project.org/web/packages/stplanr) Limited routing functions based on dodgr and igraph, and some other services.
* [gtfsrouter](https://cran.r-project.org/web/packages/gtfsrouter/index.html) For integrating GTFS public transport timetables

## Packages on GitHub

* [Open Route Service](https://giscience.github.io/openrouteservice-r/) Connect to ORS website
* [TransportAPI](https://github.com/mem48/transportAPI) An ITS Leeds Package, in development
* [OpenTripPlanner](https://github.com/ITSLeeds/opentripplanner) An ITS Leeds Package, local or remote OTP routing
* [graphhopper](https://github.com/graphhopper/directions-api-clients/tree/master/r)


## Getting some Routes

Many services require you to sign up for a free API key, to save some time we will use the TransportAPI package that allows a small amount of routing to be done without an API key.

**NOTE** If you want to do more routing please signup for a free API key at https://developer.transportapi.com/signup 

We will install the packages

```{r, eval=FALSE}
# Install packages from GitHub
devtools::install_github("ITSleeds/transportAPI")
devtools::install_github("ITSleeds/opentripplanner") # For the bonus exercises
```

And load the packages

```{r, eval=TRUE}
# Load packages
library(sf)
library(stplanr)
library(tmap)
library(transportAPI)
library(opentripplanner)
library(pct)
library(dplyr)
```

TransportAPI you to signup for a [free API key](https://developer.transportapi.com/signup). 

```{r, eval=FALSE}
usethis::edit_r_environ()
# TRANSPORTAPI_app_id=your_id_here
# TRANSPORTAPI_app_key=your_key_here
```

Now we will get some data from the [PCT](http://www.pct.bike). The `get_pct_lines` function returns the desire lines from the PCT.

```{r, eval=TRUE}
lines <- pct::get_pct_lines("isle-of-wight", "commute","lsoa")
lines <- lines[,c("id","geo_code1","geo_code2","all",
                "bicycle","car_driver","train_tube","bus")]
centroids <- pct::get_pct_centroids("isle-of-wight", "commute","lsoa")
centroids <- centroids[,"geo_code"]
plot(lines$geometry, lwd = lines$all / mean(lines$all))
plot(centroids, col = "red", add = T)
```

Now we will take the top 3 desire lines and route them though Transport API. First, we will subset the top lines and create from and to coordinates

```{r, eval=TRUE}
lines_top <- top_n(lines, 10, all)
from <- centroids[match(lines_top$geo_code1, centroids$geo_code),]
to <- centroids[match(lines_top$geo_code2, centroids$geo_code),]
```

Then we will use the `tapi_journey_batch` to find multiple routes at once.

```{r, eval=TRUE}
routes_car <- transportAPI::tapi_journey_batch(from$geometry,
                                           to$geometry, 
                                           fromid = from$geo_code,
                                           toid = to$geo_code,
                                           apitype = "car")
tmap_mode("view")
tm_shape(routes_car) +
  tm_lines()
```

We need to join the number of commuters onto the geometry of the routes.

```{r, eval=TRUE}
st_geometry(lines_top) <- NULL
routes_car <- left_join(routes_car, lines_top, by = c("fromid" = "geo_code1", "toid" = "geo_code2"))
```

Finally, we can combine the routes into a route network

```{r, eval=TRUE}
rnet <- overline2(routes_car, attrib = c("all","bicycle","car_driver","train_tube","bus"))
tm_shape(rnet) +
  tm_lines(col = "car_driver", lwd = 3)
```

**Exercises:**

1. Get routes for the top 5 desiler lines for different modes e.g. car, public transport, bike
1. How are public transport routes different to car and bike routes?
1. Plot these routes togther on a map, where are there complemenatyr and conflicing routes?






## Local Routing With Open Trip Planner

We will repeat the analysis using a local routing tool. This tutorial is based on the [packgee vignette](https://itsleeds.github.io/opentripplanner/articles/opentripplanner.html)

First, we need some basic data.

```{r, eval=FALSE}
# Create Folders for Data
dir.create("OTP")
path_data <- file.path("OTP")
path_otp <- file.path(path_data, "otp.jar")
dir.create(file.path(path_data,"graphs")) # create a folder structure for the data
dir.create(file.path(path_data,"graphs","default"))

# Download OTP and Data
download.file(url = "https://repo1.maven.org/maven2/org/opentripplanner/otp/1.3.0/otp-1.3.0-shaded.jar", 
              destfile = path_otp, mode = "wb")
download.file("https://github.com/ITSLeeds/opentripplanner/releases/download/0.1/isle-of-wight-demo.zip", 
              destfile = file.path(path_data,"isle-of-wight-demo.zip") , mode="wb")
unzip(file.path(path_data,"isle-of-wight-demo.zip"), exdir = file.path(path_data,"graphs","default"))
unlink(file.path(path_data,"isle-of-wight-demo.zip"))
```

Now we set up the OTP

```{r, eval=FALSE}
log <- otp_build_graph(otp = path_otp, dir = path_data) 
otp_setup(otp = path_otp, dir = path_data)
otpcon <- otp_connect()
```

Next, we find the routes

```{r, eval=FALSE}
routes_driving <- otp_plan(otpcon, fromPlace = from, toPlace = to,
                           mode = "CAR")
```


