---
title: "Practical 5: Visualising transport data"
bibliography: ../tds.bib
toc: true
execute: 
  cache: true
  eval: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction

In this practical, we will build on the **routing techniques** from Practical 4 by exploring **data visualization** methods for transport analysis. By the end of this practical, you should be able to:

- **Process** real transport demand datasets  
- **Create** flow maps and isochrones  
- **Analyze** network centrality  
- **Visualize** multi-modal journeys  
- **Generate** 3D transport visualizations  

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "700px"
)
```

# Setup
Below are the libraries we will use throughout this practical:
```{r}
# Load necessary libraries
library(opentripplanner)  # Routing engine (OpenTripPlanner client)
library(sf)               # Spatial data handling
library(tmap)             # Thematic mapping
library(stplanr)          # Transport data functions
library(dplyr)            # Data manipulation
library(osmextract)       # OSM data handling
library(dodgr)            # Network analysis
library(rayshader)        # 3D data visualization
library(elevatr)          # Elevation data
library(gifski)           # Creating GIF animations
library(rgl)              # 3D visualization device

# Set interactive mapping mode
tmap_mode("view")
```

# Flow Map Visualization

Flow maps are useful for understanding the volume of travel between origins and destinations. In this section, we will:

1. Load desire lines (flows) data from a GeoJSON file.
2. Visualize these lines with widths or colors proportional to demand.
3. Optionally aggregate route geometries for more realistic depiction of flows along an actual road network.

```{r}
# Load Demand Data
desire_lines = read_sf("https://github.com/ITSLeeds/TDS/releases/download/22/NTEM_flow.geojson") |>
  select(from, to, all, walk, drive, cycle)

# Quick map to see the distribution of car trips
tm_shape(desire_lines) +
  tm_lines(
    lwd = "drive",          # Line width proportional to 'drive' volume
    scale = 8,              
    palette = "viridis",    
    style = "fisher",
    title.lwd = "Car Trips (NTEM Model)"
  ) +
  tm_layout(legend.bg.color = "white")
```

# Aggregating Flows with Actual Routes

Rather than drawing direct origin-destination lines, we can route each flow along the road network and then aggregate them to see which streets carry the most traffic. This uses stplanr::overline() to merge lines that overlap.

```{r}
# Download pre-routed lines for demonstration
u = "https://github.com/ITSLeeds/TDS/releases/download/22/routes_drive_25.geojson"
routes_drive = read_sf(u)

# Inspect the summary of the drive.x variable (car trips)
summary(routes_drive$drive.x)

# Optional: remove negative values if they exist (invalid or placeholder data)
routes_drive = routes_drive |> 
  mutate(drive.x = ifelse(drive.x < 0, 0, drive.x))

# Aggregate flows where routes overlap
rnet_drive = stplanr::overline(
  routes_drive, 
  attrib = "drive.x", # The attribute to sum or aggregate
  fun = sum
)

# Visualize the aggregated route network
tm_shape(rnet_drive) +
  tm_lines(
    lwd = "drive.x", 
    col = "drive.x",
    style = "jenks",       # Good for skewed data
    n = 7,
    palette = "inferno",   # 'inferno' color palette
    title.col = "AADT Equivalent",
    scale = 5
  ) +
  tm_scalebar(position = c("left", "bottom"))
```

# Temporal Visualisation 
#| TODO: the results of this code chunk are not correct. The code needs to be fixed.

Transport accessibility can change over time due to congestion, transit schedules, etc. In this section, we’ll:

1. Connect to an OpenTripPlanner (OTP) server
2. Generate isochrones for different departure times
3. Save maps for each time step, and
4. Combine those images into an animated GIF.

```{r}
# Connect to the OTP server
ip = "otp.robinlovelace.net"  # Example server - replace if needed
otpcon = otp_connect(
  hostname = ip,
  ssl = TRUE,
  port = 443,
  router = "west-yorkshire"
)
cat("Router:", paste0("https://", ip, ":443/otp/routers/", otpcon$router), "exists\n")

# Create a time sequence (e.g. 7AM to 9AM in 30-min increments)
time_sequence = seq(7, 9, by = 0.5)

# Define a function to get isochrones for a given hour and return a tmap object
animate_isochrones = function(hour) {
  # Construct a POSIXct datetime object (adjust date/time zone as needed)
  dt = as.POSIXct(
    paste0("2023-03-15 ", hour, ":00:00"), 
    tz = "Europe/London"
  )
  
  iso = otp_isochrone(
    otpcon,
    fromPlace = c(-1.558, 53.807),  # Example coordinates
    mode = "BICYCLE",
    date_time = dt,
    cutoffSec = seq(900, 3600, 900)  # 15, 30, 45, 60 minutes
  )
  
  # Create a tmap visualization of the isochrones
  tm_shape(iso) +
    tm_fill("time", alpha = 0.6, title = "Travel Time (secs)") +
    tm_borders() +
    tm_layout(
      title = paste("Isochrone at", format(dt, "%H:%M")),
      legend.outside = TRUE
    )
}

# Directory to store temporary images
tmp_img_dir = tempdir()
img_files = character(length(time_sequence))

for (i in seq_along(time_sequence)) {
  # Create the map
  plot_obj = animate_isochrones(time_sequence[i])
  
  # Build a filename for each frame
  img_file = file.path(tmp_img_dir, sprintf("frame_%03d.png", i))
  
  # Save map as PNG
  tmap_save(plot_obj, filename = img_file, width = 800, height = 600)
  img_files[i] = img_file
  
  cat("Saved", img_file, "\n")
}

# Combine PNG frames into a GIF
gifski(
  png_files = img_files,
  gif_file = "morning_peak.gif",
  width = 800,
  height = 600,
  delay = 1
)

```

# Network Centrality Analysis

Betweenness centrality is a measure of how often a particular segment (or node) lies on the shortest path between other points in the network. Segments with high betweenness are often considered critical for overall connectivity.

Below, we:

1. Download roads from OpenStreetMap using osmextract.
2. Weight the network for motor vehicle usage.
3. Compute centrality using dodgr.
4. Convert the resulting graph back to an sf object.
5. Map the centrality values.

```{r}
# Get Isle of Wight road network
# We choose 'primary', 'secondary', 'tertiary' roads for demonstration
roads = oe_get("Isle of Wight", 
                extra_tags = c("maxspeed", "oneway")) |>
  filter(highway %in% c("primary", "secondary", "tertiary"))

# Weight the street network for motorcar usage
graph = weight_streetnet(
  roads,
  wt_profile = "motorcar",
  type_col = "highway",
  id_col = "osm_id",
  keep_cols = c("maxspeed", "oneway")
)

# Calculate betweenness centrality
centrality = dodgr_centrality(graph)

# Convert to sf for visualization
centrality_sf = dodgr_to_sf(centrality)

# Visualize critical links
tm_shape(centrality_sf) +
  tm_lines(
    col = "centrality", 
    lwd = 3, 
    style = "fisher", 
    palette = "-viridis",
    title.col = "Betweenness Centrality"
  ) +
  tm_layout(
    title = "Road Network Criticality Analysis",
    frame = FALSE
  )

```


# 3D Flow Visualization

Sometimes, a 3D perspective can reveal relationships between travel flows and the underlying topography. Below, we demonstrate how to fetch elevation data and render a 3D hillshade with rayshader. We then outline how you might overlay flow lines onto a 3D terrain model.

```{r}
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))
curl::has_internet()

# Example: Elevation data near a location in the UK
coords = data.frame(x = -2.087918, y = 53.71534)
coords_sf = st_as_sf(coords, coords = c("x", "y"), crs = 4326)

# Get an elevation raster at zoom level 11 (~ 10m resolution, depending on region)
elevation = elevatr::get_elev_raster(
  locations = coords_sf,
  z = 11
)

# Convert the raster to a matrix for rayshader
elev_matrix = rayshader::raster_to_matrix(elevation)

# Create a hillshade layer
hillshade_matrix = rayshader::ray_shade(elev_matrix, zscale = 15)

# Clear existing rgl device
rgl::rgl.clear()

# Render a 3D plot of the terrain
rayshader::plot_3d(
  heightmap = elev_matrix,
  hillshade = hillshade_matrix,
  zscale = 15,
  windowsize = c(1000, 800)
)

# Adjust camera view
rgl::view3d(theta = 30, phi = 30, zoom = 0.75)
rgl::rglwidget()
```


# Conclusions

In this practical, you learned how to:

1. Load and visualize flow data to understand travel demand.
2. Aggregate flows along actual routes for more realistic flow mapping.
3. Generate isochrones at different times, illustrating how accessibility changes.
4. Analyze network centrality, identifying critical road links.
5. Experiment with 3D visualizations to integrate topography into transport analysis.