---
title: "Practical 5: Visualising transport data"
bibliography: ../tds.bib
toc: true
execute: 
  cache: true
  eval: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction

In this practical, we will build on the **routing techniques** from Practical 4 by exploring **data visualization** methods for transport analysis. By the end of this practical, you should be able to:

- Load and preprocess OD flow data
- Visualize OD lines and proportional symbol maps
- Compare walking, driving, and cycling flows
- Aggregate flows along the road network
- Identify critical road segments via network centrality

# Setup
Below are the libraries we will use throughout this practical:
```{r}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "700px"
)

# Load necessary libraries
library(opentripplanner)  # Routing engine (OpenTripPlanner client)
library(sf)               # Spatial data handling
library(tmap)             # Thematic mapping
library(stplanr)          # Transport data functions
library(dplyr)            # Data manipulation
library(osmextract)       # OSM data handling
library(dodgr)            # Network analysis

# Set interactive mapping mode
tmap_mode("view")
```

# Flow Map Visualization

Flow maps are useful for understanding the volume of travel between origins and destinations. In this section, we will:

1. Load desire lines (flows) data from a GeoJSON file.
2. Visualize these lines with widths or colors proportional to demand.
3. Optionally aggregate route geometries for more realistic depiction of flows along an actual road network.

```{r}
# Load Demand Data
desire_lines = read_sf("https://github.com/ITSLeeds/TDS/releases/download/22/NTEM_flow.geojson") |>
  select(from, to, all, walk, drive, cycle)

# Quick map to see the distribution of car trips
tm_shape(desire_lines) +
  tm_lines(
    lwd = "drive",          # Line width proportional to 'drive' volume
    scale = 8,              
    palette = "viridis",    
    style = "fisher",
    title.lwd = "Car Trips (NTEM Model)"
  ) +
  tm_layout(legend.bg.color = "white")
```

# Proportional Symbol Flow Maps
Now, let's illustrate an alternative method: proportional symbols at origin or destination points. This is useful when you want to quickly see where demand is concentrated.

```{r}
# Summarize total flows by origin

origin_flows = desire_lines |>
  group_by(from) |>
  summarise(
    total_drive = sum(drive, na.rm = TRUE),
    total_walk  = sum(walk, na.rm = TRUE),
    total_cycle = sum(cycle, na.rm = TRUE),
    geometry = st_centroid(st_union(geometry))  
  )

# Simple map with proportional circles for drive volumes

tm_shape(origin_flows) +
  tm_bubbles(
    size    = "total_drive",       # bubble size ~ drive volume
    col     = "total_drive",       # color also ~ drive volume
    border.col = "black", border.alpha = 0.5,
    palette = "Reds", style = "jenks",
    scale = 2, 
    title.col = "Drive Volume",
    title.size = "Drive Volume"
  ) +
  tm_layout(
    main.title = "Proportional Symbol Map of Drive Demand by Origin",
    legend.outside = TRUE
  )
```

Each origin is represented by a circle whose radius and color intensity reflect the total number of driving trips. You can modify palettes, breaks, and scaling to highlight variations.

# Mode-Specific Analysis

We have have columns walk, drive, cycle in desire_lines. We can map them separately or side-by-side. We can also color lines by the dominant mode.

```{r}
# Let's create 3 separate maps: drive, walk, cycle
m_drive = tm_shape(desire_lines) +
  tm_lines(
    lwd = "drive", 
    col = "drive",
    title.lwd = "Drive Vol",
    palette = "Reds",
    style = "quantile"
  ) +
  tm_layout(main.title="Driving Flows")

m_walk = tm_shape(desire_lines) +
  tm_lines(
    lwd = "walk", 
    col = "walk",
    title.lwd = "Walk Vol",
    palette = "Greens",
    style = "quantile"
  ) +
  tm_layout(main.title="Walking Flows")

m_cycle = tm_shape(desire_lines) +
  tm_lines(
    lwd = "cycle", 
    col = "cycle",
    title.lwd = "Cycle Vol",
    palette = "Blues",
    style = "quantile"
  ) +
  tm_layout(main.title="Cycling Flows")

tmap_arrange(m_drive, m_walk, m_cycle, ncol=3)

```

This tmap_arrange() will output a single figure with three columns, each illustrating flows by one mode. Students can visually compare the differences: maybe driving flows are much thicker on longer corridors, while walking flows are concentrated in the city center.



# Aggregating Flows with Actual Routes

Rather than drawing direct origin-destination lines, we can route each flow along the road network and then aggregate them to see which streets carry the most traffic. This uses stplanr::overline() to merge lines that overlap.

```{r}
# Download pre-routed lines for demonstration
u = "https://github.com/ITSLeeds/TDS/releases/download/22/routes_drive_25.geojson"
routes_drive = read_sf(u)

# Inspect the summary of the drive.x variable (car trips)
summary(routes_drive$drive.x)

tm_shape(routes_drive) +
  tm_lines(
    lwd = "drive.x",
    col = "drive.x",
    palette = "inferno",
    style = "fisher",
    title.lwd = "Drive Flow",
    title.col = "Drive Flow"
  ) +
  tm_layout(main.title="Road Congestion (drive_total)")
```

# Network Centrality Analysis

Betweenness centrality indicates how often a road (or node) lies on the shortest path between other points in a network. Roads with high centrality are typically crucial for overall connectivity.

Here, we demonstrate how to:

1. Download roads from OpenStreetMap (OSM) for the Isle of Wight.
2. Weight the network for motor vehicle usage.
3. Compute betweenness centrality with dodgr.
4. Convert the results back to sf for mapping.

```{r}
# Get Isle of Wight road network
# We choose 'primary', 'secondary', 'tertiary' roads for demonstration
roads = oe_get("Isle of Wight", 
                extra_tags = c("maxspeed", "oneway")) |>
  filter(highway %in% c("primary", "secondary", "tertiary"))

# Weight the street network for motorcar usage
graph = weight_streetnet(
  roads,
  wt_profile = "motorcar",
  type_col = "highway",
  id_col = "osm_id",
  keep_cols = c("maxspeed", "oneway")
)

# Calculate betweenness centrality
centrality = dodgr_centrality(graph)

# Convert to sf for visualization
centrality_sf = dodgr_to_sf(centrality)

# Visualize critical links
tm_shape(centrality_sf) +
  tm_lines(
    col = "centrality", 
    lwd = 3, 
    style = "fisher", 
    palette = "-viridis",
    title.col = "Betweenness Centrality"
  ) +
  tm_layout(
    title = "Road Network Criticality Analysis",
    frame = FALSE
  )

```

High values in the centrality column indicate roads that act as vital connectors in the regional transport network.


# Extra Exercises: 3D Visualization

A 3D perspective can often reveal relationships between travel flows and the underlying topography more effectively. Below, we demonstrate how to retrieve elevation data and render a 3D hillshade using the rayshader package. You may also be interested in overlaying flow lines onto a 3D terrain model to enhance visualization.

Note: the following code requires you to install the `rayshader` `elevatr` `gifski` `rgl` package. Results not shown in website.

```{r}  
library(rayshader)        # 3D data visualization
library(elevatr)          # Elevation data
library(gifski)           # Creating GIF animations
library(rgl)              # 3D visualization device

assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))
curl::has_internet()
# Example: Elevation data near a location in the UK
coords = data.frame(x = -2.087918, y = 53.71534)
coords_sf = st_as_sf(coords, coords = c("x", "y"), crs = 4326)
# Get an elevation raster at zoom level 11 (~ 10m resolution, depending on region)
elevation = elevatr::get_elev_raster(
  locations = coords_sf,
  z = 11
)
# Convert the raster to a matrix for rayshader
elev_matrix = rayshader::raster_to_matrix(elevation)
# Create a hillshade layer
hillshade_matrix = rayshader::ray_shade(elev_matrix, zscale = 15)
# Clear existing rgl device
rgl::rgl.clear()
# Render a 3D plot of the terrain
rayshader::plot_3d(
  heightmap = elev_matrix,
  hillshade = hillshade_matrix,
  zscale = 15,
  windowsize = c(1000, 800)
)
# Adjust camera view
rgl::view3d(theta = 30, phi = 30, zoom = 0.75)
rgl::rglwidget()

```

# Conclusions

In this practical, you learned how to:

1. Create flow maps to visualize travel demand from an OD dataset.
2. Compare flows by mode (driving, walking, cycling) to understand differences in spatial patterns.
3. Aggregate routes along the road network (using stplanr::overline) to highlight heavily used corridors.
4. Compute betweenness centrality (using dodgr) to pinpoint critical road segments crucial for connectivity.
