---
format: pdf
# Try replacing the above with this for PDF output
# format: pdf
title: "The impact of lockdowns on collisions"
number-sections: true
bibliography: references.bib
execute: 
  cache: refresh
---

# Introduction

This document explores the impact of lockdowns on the rate of collisions.

# Data and methods

Datasets were taken from the STATS19 database and processed with the `stats19` R package [@lovelace2019].

# Results

```{r}
#| include: false
library(tidyverse)
library(stats19)
# dl_stats19(year = 2020, type = "collision")
collisions_2020 = get_stats19(year = 2020, type = "collision")
```

See the results of downloading national crash data from 2020 in @fig-crashes-per-day.

```{r}
#| label: fig-crashes-per-day
#| echo: false
#| fig-cap: "Crashes over time in 2020"
collisions_per_day = collisions_2020 |> 
  group_by(date) |> 
  summarise(
    n = n()
  )
collisions_per_day |> 
  ggplot(aes(date, n)) +
  geom_line(colour = "red") 
```

The equivalent trend for Leeds is shown in @fig-trend-leeds.

The column names available to us are:

```{r}
names(collisions_2020)
```


```{r}
#| echo: false
#| label: fig-trend-leeds
# Check the local authority of crashes:
collisions_wy = collisions_2020 |> 
  filter(police_force == "West Yorkshire")
collisions_per_day_wy = collisions_wy |> 
  group_by(date) |> 
  summarise(
    n = n()
  )
collisions_per_day_wy |> 
  ggplot(aes(date, n)) +
  geom_line(colour = "red") 


```

Let's put them side-by-side.

```{r}
# collisions_per_day_combined = bind_rows(
#   collisions_per_day |> mutate(Area = "GB"),
#   collisions_per_day_wy |> mutate(Area = "West Yorksire")
# )
# # Not sure why the scales here do not work:
# collisions_per_day_combined |> 
#   ggplot(aes(date, n)) +
#   geom_line() +
#   facet_grid(~Area, shrink = TRUE)
g1 = collisions_per_day |> 
  ggplot(aes(date, n)) +
  geom_line(colour = "red") 
g2 = collisions_per_day_wy |> 
  ggplot(aes(date, n)) +
  geom_line(colour = "red") 
library(patchwork)
```

```{r}
#| echo: false
#| layout-ncol: 2
g1
g2
```


# Conclusion

# References
