---
format: pdf
title: "Transport Data Science Coursework Plan"
bibliography: 2-pager.bib
---

```{=html}
<!-- 

- Deadline for non-assessed submission of a .zip file containing 2 page pdf document with ideas before the final submission. The document will allow you to ask questions (e.g. "does this sound like a reasonable input dataset and topic?") and describe progress on reading in input datasets and the analysis plan. The document will contain:
  - A draft title of your topic

  - The structure of the document could include
    - Topics considered
    - Input datasets
    - Analysis plan - I suggest creating a workflow diagram for this, e.g. as presented [here](https://user-images.githubusercontent.com/1825120/127524923-7d9f5511-84a6-430b-8de9-a603a5524f39.png)
    - Motivation for choosing this topic
    - Questions and options -->
```
# Title

The draft title of my project is Exploring the Accessibility of Parks for Wombats in Melbourne.

\[To be confirmed\]

# Research questions considered and motivations

Wombats are a key species

# Prior work

@pereira2023 provides an introduction to the field and guidance on accessibility analysis.

@lovelace2021 provides an overview of existing tools for wombat accessibility analysis.

# Input data

I will use routes from the Propensity to Cycle Tool project as the main input dataset.

@lovelace2017 describes these datasets in detail.

See <https://itsleeds.github.io/QGIS-intro/data-sources-for-transport-data.html>

And <https://beta.source.coop/>

# Questions for the course team

Is the input dataset and research question too ambitious?

I was wondering about using regression with the `xgboost` package.
I have tried it and managed to get some results shown below.

# Analysis plan

I have 4 full weeks to complete the 10 page report, in that time I will

-   Week 1: complete to introduction to the report and references, write data cleaning code and describe the input dataset

-   ...

-   ...

-   ...

# Importing the data and initial exploration

See below for an initial exploration of the data

```{r}
#| include: false
library(stplanr)
library(tidyverse)
routes_imported = routes_fast_sf
names(routes_imported)
desire_lines = flowlines_sf
desire_lines$id = 1:nrow(desire_lines)
routes_imported$id = 1:nrow(routes_imported)
routes_joined = left_join(
    routes_imported,
    desire_lines |> sf::st_drop_geometry()
)
```

We aggregated the data with the `overline()` function from the `stplanr` package, resulting in @fig-routes (right).

```{r}
#| label: fig-routes
#| layout-ncol: 2
#| fig-cap: "Routes in Melbourne, Australia"
#| fig-subcap: "Left: raw data; right: aggregated data"
routes_joined |>
  select(Bicycle) |>
  plot()
routes_overline = overline(routes_joined, "Bicycle")
plot(routes_overline)
```

# References