---
title: "stats19-locations"
# output: bookdown::pdf_document2
bibliography:
  - saferactive.bib
  - my-bib.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
# load packages
library(stats19)
library(tidyverse)
library(tmap)
```

# Introduction

I suggest taking a look at and perhaps citing these papers and reports:

[@aldred_cycling_2018]

[@ding_roles_2020]

[@feleke_comparative_2018]

[@grolemund_data_2016]

See here for more info: https://www.zotero.org/groups/418217/energy-and-transport/collections/9PTF9F44/item-list

<!-- [@mulvaney_cycling_2015] -->

See here and in the .bib files for more citations:




# Descriptive data analysis


## Case study of Leeds-Bradford Cycle Superhighway

A case study of a well-known cycleway in Leeds is shown in Figure \@ref(fig:cycleway1).

```{r, eval=TRUE, fig.cap="Illustration of the Leeds-Bradford cycle superhighway. Source: OpenStreetMap.", fig.width=9}
# read in saved dataset
u = "https://github.com/ropensci/stats19/releases/download/1.1.0/superhighway.Rds"
superhighway = readRDS(url(u))
plot(superhighway$geometry)
# tm_shape(superhighway) +
#   tm_lines("highway") +
#   tm_layout(asp = 5, frame = FALSE)
# +
#   tm_layout(legend.outside = TRUE)
```


## Cycleway data

## Crash data

## Cycling behaviour data

[@lovelace_propensity_2017]

## Explanatory variables

E.g. speed from OSM data and number of junctions

presence/absence of cycleway

# Modelling crash rates on cycleways


# Discussion

## Limitations

## Policy relevance

## Further research


# Conclusions

```{r setup, eval=TRUE}
crashes = get_stats19(2012:2018, output_format = "sf")
#2018
# fail use map_dfr...
# challenge: get more years
casualties_2018 = get_stats19(year = 2018, type = "cas")
casualties_cyclist_2018 = casualties_2018 %>% 
  filter(casualty_type == "Cyclist")
crashes_cyclist_2018 = crashes %>% 
  filter(accident_index %in% casualties_cyclist_2018$accident_index)
crashes_cyclist_2018 = sf::st_transform(crashes_cyclist_2018, 4326)
crashes_near_superhighway_2018 = crashes_cyclist_2018[superhighway_buffer, ]

#total
casualties = get_stats19(2012:2018, type = "cas") 
casualties_cyclist = casualties %>% 
  filter(casualty_type == "Cyclist")
crashes_cyclist = crashes %>% 
  filter(accident_index %in% casualties_cyclist$accident_index)
```

## OpenStreetMap data

```{r}
library(sf)
library(osmdata)
data_osm = opq("leeds") %>% 
  add_osm_feature(key = "name", value = "Cycle Superhighway 1") %>% 
  osmdata_sf()
data_osm
plot(data_osm$osm_lines$geometry)
wy = pct::get_pct_zones("west-yorkshire")
superhighway = data_osm$osm_lines[wy, ]
mapview::mapview(superhighway)
saveRDS(superhighway, "superhighway.Rds")
piggyback::pb_upload("superhighway.Rds", repo = "ropensci/stats19")
piggyback::pb_download_url("superhighway.Rds", repo = "ropensci/stats19")
```

```{r}
# get buffer
# the distance here use 20 which will not include other routes data but won't miss the junction data.  
superhighway_buffer = stplanr::geo_buffer(superhighway, dist = 20) %>% 
  sf::st_union()
mapview::mapview(superhighway_buffer)
crashes_cyclist = sf::st_transform(crashes_cyclist, 4326)
crashes_near_superhighway = crashes_cyclist[superhighway_buffer, ]
mapview::mapview(crashes_near_superhighway)
summary(crashes_near_superhighway$datetime)
table(crashes_near_superhighway$accident_severity)
```

## Questions

1. How to get casualties for more years, e.g. 2012 to 2018 (not just 2018)?

I am not quite sure why the script code "casualties = get_stats19(2012:2018, type = "cas")" doesn't works. I run this code and get 56 observations in the end. "map_dfr" seems return data frames created by row-binding and column-binding respectively, so I first thought it may fail because different variable number (some years have 15 variables and some 16). But it turns out the NA will fill in the lack variable.


2. How to estimate the risk per year?
This raises another question, how to estimate the exposure near the cycleway?

To avoid the "safety in number", the cyclist flow should be considered. On different segments of the road. The bicycle number varys from 0 to 387, it leads the exposure data varys. So it is better to consider separately for each crash location than treat the superhighway equally risky.

Because the speed limit does not differ a lot on the entire road, the distance is the other factor of exposure. The exposure equation is multiply cycling volume per yearby road segment length (in vehicle-kilometers, VKM). The risk equals the crash number per year divide by exposure. The crash number could separates KSI from slight injuries.

```{r}
#exposure data of the crash point
rnet_wy_cycling = pct::get_pct_rnet(region = "west-yorkshire")
plot(rnet_wy_cycling$geometry)
rnet_cycleway = sf::st_intersection(rnet_wy_cycling, superhighway_buffer)
plot(rnet_cycleway)
mapview::mapview(rnet_cycleway)

#exposure of the crash point
test1<-crashes_near_superhighway
test2<-rnet_cycleway

#if the crash geometry located in the segments.
mat = st_intersects(test1$geometry, test2$geometry, sparse = FALSE)
apply(mat, 1, any)
#all in false maybe because of the geometry deviation (not exactly located). 
#alter to st_distance function. It calculates the distance matrix in meters using Great Circle distance.
#treat the crashes happen in the closed segement. 
#The most minimum distances are controlled under 50m and the biggest one is 105m. 
d<-data.frame(st_distance(test1$geometry,test2$geometry))
nn.dist <- apply(d, 1, function(x) {
  return(sort(x, partial = 1)[1])
})
nn.index <- apply(d, 1, function(x) {order(x, decreasing=F)[1]})
newdata <- cbind(test1, test2[nn.index,], apply(d, 1, function(x) sort(x, decreasing=F)[2]))

newdata$length_km =
  as.numeric(sf::st_length(newdata$geometry.1))
newdata$km_cycle_peryear = newdata$length_km * newdata$bicycle*2*260

nrow(newdata) /
  (sum(newdata$km_cycle_peryear) / 1e9)

```



How to estimate how numbers cycling changes per year (difficult, not essential to answer)?

```{r}
rnet_wy_cycling = pct::get_pct_rnet(region = "west-yorkshire")
plot(rnet_wy_cycling$geometry)
rnet_cycleway = sf::st_intersection(rnet_wy_cycling, superhighway_buffer)
plot(rnet_cycleway)
mapview::mapview(rnet_cycleway)
# how many cycle journey segments in total?
sum(rnet_cycleway$bicycle)
# estimate million km cycled per year...
rnet_cycleway$length_km =
  as.numeric(sf::st_length(rnet_cycleway)) / 1000
summary(rnet_cycleway$length_km)
# estimate km cycled per working morning
rnet_cycleway$km_per_morning_workday =
  rnet_cycleway$length_km * rnet_cycleway$bicycle
sum(rnet_cycleway$km_per_morning_workday)
rnet_cycleway$km_per_year =
  rnet_cycleway$km_per_morning_workday * 2 * 200
sum(rnet_cycleway$km_per_year)
# estimate risk - total crashes per bkm
nrow(crashes_near_superhighway) /
  (sum(rnet_cycleway$km_per_year) / 1e9)
# estimate risk - total crashes per bkm
1 /
  (sum(rnet_cycleway$km_per_year) / 1e9)
```

```{r}
remotes::install_github("itsleeds/osmextractr")
library(osmextractr)
wy_osm = oe_get(place = "West Yorkshire", extra_attributes = c("maxspeed"))
pryr::object_size(wy_osm)
wy_osm_cycleway = wy_osm[superhighway_buffer, , op = sf::st_within]
mapview::mapview(wy_osm_cycleway["highway"])
wy_osm_cycleway$speed = as.character(wy_osm_cycleway$maxspeed)
table(wy_osm_cycleway$speed)
wy_osm_cycleway$speed_numeric = as.numeric(gsub(pattern = " mph", replacement = "", wy_osm_cycleway$maxspeed))
table(wy_osm_cycleway$speed_numeric)
mean(wy_osm_cycleway$speed_numeric, na.rm = TRUE) # the average speed limit on the cycleway
table(wy_osm_cycleway$highway)
# calculate percentage on primary:
sum(wy_osm_cycleway$highway == "primary", na.rm = TRUE) / 
  sum(!is.na(wy_osm_cycleway$highway))
# 0.1034483
# calculate percentage on primary:
sum(wy_osm_cycleway$highway == "secondary", na.rm = TRUE) / 
  sum(!is.na(wy_osm_cycleway$highway))
```


# References

