---
title: "4 hour workshop: R for Transport Data Analysis"
output: github_document
bibliography:
  - ../tds.bib
  - ../software.bib
---

Transport Data Science course, based on materials hosted at https://github.com/ITSLeeds/TDS

Home of this course: git.io/tds4hr

Assumed prior knowledge:

- Working knowledge of R, e.g. have completed: 
  - Introduction to R free DataCamp course: https://www.datacamp.com/courses/free-introduction-to-r
  - Recommended reading: Section 4.2 of *Efficient R Programming* [@gillespie_efficient_2016]: https://csgillespie.github.io/efficientR/efficient-workflow.html#package-selection

- Some knowledge of tidyverse is highly recommended, e.g. already know the contents of or have read-up on **and tried examples from** the following 'cheatsheets':
  - Tidyverse for Beginners DataCamp Cheat Sheet (see [here](https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf)). **At a minimum you will have completed up to Chapter 5 on this online course** (this may take ~2 hours)
  - Data wrangling with dplyr and tidyr RStudio cheet cheet (see [here](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf))
  
<!-- **It's in the Analysis directorate** -->

## Computing Requirements

- Desktops with recent versions of R and RStudio installed, including:
  - R 3.6.0 or later recommended (although any R 3.5 release should work): **download and install the latest version from** https://cran.r-project.org/
  - A recent version of RStudio: **download and install the latest version from** https://www.rstudio.com/products/rstudio/download/
  - Up-to-date versions of **tidyverse** and **sf** packages, at a minimum
  - Other packages we'll use are shown in the code below, which installs them
  
The packages used can be installed as follows:

```{r, eval=FALSE}
install.packages("remotes")
pkgs = c(
  "osmdata",
  "pct",
  "sf",
  "stats19",
  "stplanr",
  "tidyverse",
  "tmap"
)
remotes::install_cran(pkgs)
```

- Own laptops should be used, unless the necessary packages work on the cluster

- Data: all data will be either provided by packages or downloaded on the day (assuming good internet)
  - A test to check if data downloads work is accessing stats19 data (check this works):

```{r, eval=FALSE}
crashes = stats19::get_stats19(year = 2017, type = "accidents", ask = FALSE)
```

- To test you can download data from OSM, try the following code:

```{r}
# set-up
library(osmdata)
library(tmap)
ttm()

# get data
d = opq("madrid") %>% 
  add_osm_feature("highway", "cycleway") %>% 
  osmdata_sf()

# plot data
tm_shape(d$osm_lines) + 
  tm_lines(col = "blue", lwd = 5) +
  tm_view(basemaps = leaflet::providers$OpenStreetMap.BlackAndWhite)
```


## Venue and course times

The course will take place from 10:00 to 14:00 at [Edificio Multiusos Ciudad Universitaria](https://www.google.com/maps/place/Edificio+Multiusos/@40.4500279,-3.7331934,15z/data=!4m5!3m4!1s0x0:0x721fc321c370faad!8m2!3d40.4500279!4d-3.7331934), 28040 Madrid

See https://forms.gle/m5zraC8QQH5AydgT8

<!-- - Download data from open roads: put on test scripts -->
  
## Required reading

The course will build on Chapter 12 of *Geocomputation with R* [@lovelace_geocomputation_2019]: http://geocompr.robinlovelace.net/transport.html

## Optional extra reading

- Optional extra reading: for context, you may want to read-up on:
  - Overview of GIS in R, e.g. in Chapter 1 of *Geocomputation with R* or this blog post: https://www.jessesadler.com/post/gis-with-r-intro/
  - stplanr: A package for transport planning [@lovelace_stplanr:_2018]
  - R for data science [@grolemund_r_2016]

### Agenda

- 10:00 - 11:00: Introduction and getting started with R

- 11:00 - 12:00: Getting and analysing transport data: examples with `stats19`, `pct` and `osmdata` packages

12:00 - 12:30: Break

- 12:30 - 13:00: Origin-destination (OD) data analysis with `stplanr`

- 13:00 - 14:00: From routes to route networks and data and methods for assessing cycling potential

## Exercises

## Getting started

See section 1 of the exercises: https://git.io/tds2dayex


```{r, out.width="30%", fig.show='hold', message=FALSE}
d = data.frame(x = 1:9, y = (1:9)^2)
plot(d)           # base R
library(ggplot2)  # with a package
ggplot(d) +
  geom_point(aes(x, y))
```

## Getting tranport data

Starting point: run the following commands.

```{r}
library(tidyverse)
library(stats19)
library(pct)
regions = pct::pct_regions
iow = regions %>% 
  filter(region_name == "isle-of-wight")
plot(iow)
vehicles = get_stats19(year = 2017, type = "vehicles", ask = FALSE)
```



1. Download and plot all crashes reported in Great Britain in 2017 (hint: see [the stats19 vignette](https://cran.r-project.org/web/packages/stats19/vignettes/stats19.html))
1. Find the function in the `stats19` package that converts a `data.frame` object into an `sf` data frame. Use this function to convert the road crashes into an `sf` object, called `crashes_sf`, for example.
1. Filter crashes that happened in the Isle of Wight based on attribute data (hint: the relevant column contains the word `local`)
1. Filter crashes happened in the Isle of Wight using geographic subsetting (hint: remember `st_crs()`?)

## Origin-destination data with stplanr

The starting point of these exercises is to work with the Origin-Destination data in the `stplanr` package.

```{r}
library(stplanr)
head(flow)
plot(flow$On.foot, flow$Bicycle)
cor(flow$On.foot, flow$Bicycle)
plot(zones_sf)
l = od2line(flow, zones_sf)
plot(l)
```

- Plot the map interactively using mapview
- Highlight the desire line along which most people walk

You can analyse OD data in more detail as follows.

```{r}
l1 = l %>% 
  top_n(1,  Bicycle)
plot(l1)
```





### Exercises

1. Create an object representing desire lines in Isle of Wight, e.g, with: `desire_lines_all = pct::get_pct_lines(region = "isle-of-wight")`
1. Subset the desire lines with a value of `all` of 200 or above
1. Create a buffer of 500 m for each desire line and calculate the number of crashes that happened within each (using STATS19 data downloaded in the previous exercise)
1. Create a faceted plot showing the temporal distribution of crashes in Isle of Wight (you can choose whether to show these over the months of the year, over days of the week, or over the hours of a day)
1. Do a spatiotemporal subset to identify the crashes that happened within the most commonly travelled desire line between 07:00 and 10:00 during weekdays.


```{r, eval=FALSE}

u = "https://www.openstreetmap.org/trace/2992569/data"
download.file(u, "track.gpx")
sf::st_layers("track.gpx")

tracks = sf::st_read("track.gpx", layer= "track_points")
# ?st_read
mapview::mapview(tracks)


library(stplanr)
library(tidyverse)
f = flow[1:3, 1:5]
f_tidy = flow %>% 
  select(1:5) %>% 
  slice(1:3)
f
class(f)
z = zones_sf
l = od2line(f, z)
plot(l[2, ])
nrow(l)
r = line2route(l, route_fun = route_osrm)
plot(r)

# get data from PCT
od = pct::get_od()
od
class(od)
iow = pct::get_pct_centroids("isle-of-wight")
plot(iow)
mapview::mapview(iow)

od_iow = od %>%
  filter(geo_code1 %in% iow$geo_code) %>% 
  filter(geo_code2 %in% iow$geo_code) 

l_iow = od2line(od_iow, iow)
mapview::mapview(l_iow)

l_iow = l_iow %>% 
  mutate(percent_drive = car_driver / all * 100) %>% 
  mutate(distance_euclidean = sf::st_length(l_iow)) %>% 
  mutate(distance_euclidean = as.numeric(distance_euclidean))
l_iow$percent_drive
class(l_iow$distance_euclidean)

summary(l_iow$percent_drive)
summary(l_iow$distance_euclidean)

l_iow_car_dependent = l_iow %>% 
  filter(percent_drive > 75) %>% 
  filter(distance_euclidean < 10000)

mapview::mapview(l_iow_car_dependent)

r_fast = line2route(l_iow_car_dependent)
r = line2route(l_iow_car_dependent, plan = "quietest")

mapview::mapview(r) +
  mapview::mapview(r_fast)

r_fast$bicycle = l_iow_car_dependent$bicycle

iow_rnet = overline(r_fast, "bicycle")
plot(iow_rnet)

# git.io/tds2dayex

# travel patterns in/around complutense
library(osmdata)
library(tidyverse)
complu = opq(bbox = "Madrid") %>% 
  add_osm_feature(key = "name", value = "Complutense", value_exact = FALSE) %>% 
  osmdata_sf()

complu

complu_poly = complu$osm_polygons
mapview::mapview(complu_poly)
complu1 = complu_poly %>% 
  mutate(area = as.numeric(sf::st_area(.))) %>% 
  top_n(1, area)
mapview::mapview(complu1)

# get 1 km buffer
complu_buffer = complu1 %>% 
  st_transform(3087) %>% 
  st_buffer(1000) %>% 
  st_transform(4326)

mapview::mapview(complu_buffer)
b = sf::st_bbox(complu_buffer)

# s = dodgr::dodgr_streetnet(bbox = b)
osm = opq(bbox = b) %>% 
  add_osm_feature("highway") %>% 
  osmdata_sf()

mapview::mapview(osm$osm_lines)

rnet = SpatialLinesNetwork(osm$osm_lines)

rnet@sl$centrality = igraph::edge_betweenness(rnet@g)

plot(rnet@sl["centrality"])

library(tmap)
ttm()
tm_shape(rnet@sl) +
  tm_lines(lwd = "centrality", scale = 9, col = "highway")

# try cyclstreets in Madrid

spain = rnaturalearth::ne_(scale = 10, country = "Spain", returnclass = "sf")
names(spain)

sf::st_geometry_type(spain)

sf::write_sf(spain, "/tmp/spain.gpkg")

u = "http://www.ine.es/censos2011_datos/cartografia_censo2011_nacional.zip"
download.file(u, "data.zip")
unzip("data.zip")
spain = sf::read_sf("SECC_CPV_E_20111101_01_R_INE.shp")
spain_sub = spain %>% 
  filter(CUSECCUMUN == "01001")
mapview::mapview(spain[1:100, ])
```

  
## References

